---
title: "Data Exploration : Attrition Analytics "
author: "Ameya Ghasias, Dhijith CD, Urvi Kalia"
date: "23 June 2016"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
    toc_float: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The purpose of this documentation is to showcase the findings of the exploratory data analysis on attrition dataset, obtained from IBM watson analytics website <reference>.
 
To brief about the dataset , It contains details about past employees of a company (details not known).It includes  demographic details, work related details, various ratings and whether the employee has left the company or not.

Since the source of the data is not known, here are some asumptions on which further analysis will be based.

* The given data is a representive sample.It is assumed that the simple random sampling technique is used to draw/collect data and hence the inferences can be generaized.
* It is assumed that the data collection was as per required standards with minimal or no data collection issues.

## Setup

The initial setup includes loading the required packages and the data.

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(grid)
library(gridExtra)
library(gmodels)
library(corrplot)
```

### Load Data
```{r load-data}
employeeData <-read.csv("HR-Employee-Attrition.csv",header = TRUE)
sum(is.na(employeeData))
dim(employeeData)
```

```{r initial-setup,echo=FALSE}
theme <- theme(axis.text=element_text(size=7),axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),plot.title = element_text(size=9, margin = margin(10, 0, 10, 0))) 
```

The dataset contain no missing values.

### PreProcessing 

Mentioned below are some of the preprocessing done :

1. Few variables have numeric ratings.These are transformed to a readable format as factors.Attributes which are transformed as factors are :          

    + *Education*
    + *Environment Satisfaction*
    + *Job Involvement*
    + *Job Satisfaction*
    + *Performance rating*
    + *Relationship satisfaction*
    + *Work life balance*
    

2. Atrributes named *JobLevel* and *StockOptionLevel* are also changed to categorical , as they represent levels.
3. Attributes named *EmployeeCount* , *Over18* , *StandardHours* are ignored, as they are constants or should be said uniformly distributed across.
4. *EmployeeNumber* to be considered as row names.
5. *Attrition* is the response variable.

```{r transform-data,echo=FALSE}
## transformation to factors 
## Education 
employeeData$Education[employeeData$Education==1] = "Below College"
employeeData$Education[employeeData$Education==2] = "College"
employeeData$Education[employeeData$Education==3] = "Bachelor"
employeeData$Education[employeeData$Education==4] = "Master"
employeeData$Education[employeeData$Education==5] = "Doctor"
employeeData$Education <- factor(employeeData$Education)

## EnvironmentSatisfaction
employeeData$EnvironmentSatisfaction[employeeData$EnvironmentSatisfaction==1] ="Low"
employeeData$EnvironmentSatisfaction[employeeData$EnvironmentSatisfaction==2] ="Medium"
employeeData$EnvironmentSatisfaction[employeeData$EnvironmentSatisfaction==3] ="High"
employeeData$EnvironmentSatisfaction[employeeData$EnvironmentSatisfaction==4] ="Very High"
employeeData$EnvironmentSatisfaction <- factor(employeeData$EnvironmentSatisfaction)

## JobInvolvement
employeeData$JobInvolvement[employeeData$JobInvolvement==1]="Low"
employeeData$JobInvolvement[employeeData$JobInvolvement==2] ="Medium"
employeeData$JobInvolvement[employeeData$JobInvolvement==3] ="High"
employeeData$JobInvolvement[employeeData$JobInvolvement==4] ="Very High"
employeeData$JobInvolvement <- factor(employeeData$JobInvolvement)

## JobSatisfaction
employeeData$JobSatisfaction[employeeData$JobSatisfaction==1]="Low"
employeeData$JobSatisfaction[employeeData$JobSatisfaction==2] ="Medium"
employeeData$JobSatisfaction[employeeData$JobSatisfaction==3] ="High"
employeeData$JobSatisfaction[employeeData$JobSatisfaction==4] ="Very High"
employeeData$JobSatisfaction <- factor(employeeData$JobSatisfaction)

## PerformanceRating
employeeData$PerformanceRating[employeeData$PerformanceRating==1]="Low"
employeeData$PerformanceRating[employeeData$PerformanceRating==2] ="Good"
employeeData$PerformanceRating[employeeData$PerformanceRating==3] ="'Excellent"
employeeData$PerformanceRating[employeeData$PerformanceRating==4] ="Outstanding"
employeeData$PerformanceRating <- factor(employeeData$PerformanceRating)

## RelationshipSatisfaction
employeeData$RelationshipSatisfaction[employeeData$RelationshipSatisfaction==1]="Low"
employeeData$RelationshipSatisfaction[employeeData$RelationshipSatisfaction==2] ="Medium"
employeeData$RelationshipSatisfaction[employeeData$RelationshipSatisfaction==3] ="High"
employeeData$RelationshipSatisfaction[employeeData$RelationshipSatisfaction==4] ="Very High"
employeeData$RelationshipSatisfaction <- factor(employeeData$RelationshipSatisfaction)

## WorkLifeBalance
employeeData$WorkLifeBalance[employeeData$WorkLifeBalance==1]="Bad"
employeeData$WorkLifeBalance[employeeData$WorkLifeBalance==2] ="Good"
employeeData$WorkLifeBalance[employeeData$WorkLifeBalance==3] ="Better"
employeeData$WorkLifeBalance[employeeData$WorkLifeBalance==4] ="Best"
employeeData$WorkLifeBalance <- factor(employeeData$WorkLifeBalance)

## JobLevel and StockOptionLevel: categorical
employeeData$JobLevel <- factor(employeeData$JobLevel)
employeeData$StockOptionLevel <- factor(employeeData$StockOptionLevel)
employeeData <- employeeData[-c(9,21,26)]

```

After all the above preprocessing we are left with 31 variables , including the response variable.17 of them being categorical variables and 14 numeric.


##Categorical  variables 

```{r cgroup1,echo=FALSE}
department <- ggplot(employeeData,aes(Department,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Department distribution") + theme
businessTravel <- ggplot(employeeData,aes(BusinessTravel,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Business Travel distribution") + theme
grid.arrange(department,businessTravel,ncol=2)
```

```{r cgroup2,echo=FALSE}
education <- ggplot(employeeData,aes(Education,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Education distribution") + theme
educationField <- ggplot(employeeData,aes(EducationField,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Education Field distribution") + theme
grid.arrange(education,educationField,ncol=2)
```

```{r cgroup3,echo=FALSE}
environmentSatisfaction <- ggplot(employeeData,aes(EnvironmentSatisfaction,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Environment Satisfaction distribution") + theme
jobInvolvement <- ggplot(employeeData,aes(JobInvolvement,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Job Involvement distribution") + theme
grid.arrange(environmentSatisfaction,jobInvolvement,ncol=2)
```


```{r cgroup4,echo=FALSE}
gender <- ggplot(employeeData,aes(Gender,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Gender distribution") + theme
maritalStatus <- ggplot(employeeData,aes(MaritalStatus,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Marital Status distribution") + theme
grid.arrange(gender,maritalStatus,ncol=2)
```

```{r cgroup5,echo=FALSE}
jobRole <- ggplot(employeeData,aes(JobRole,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Job Role distribution") + theme
jobSatisfaction <- ggplot(employeeData,aes(JobSatisfaction,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Job Satisfaction distribution") + theme

jobLevel <- ggplot(employeeData,aes(JobLevel,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Job Level distribution") + theme
stockOptionLevel <- ggplot(employeeData,aes(StockOptionLevel,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Stock OptionLevel distribution") + theme
grid.arrange(jobRole,jobSatisfaction,ncol=2)
grid.arrange(jobLevel,stockOptionLevel,ncol=2)
```

```{r cgroup6,echo=FALSE}
overTime <- ggplot(employeeData,aes(OverTime,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Over Time distribution") + theme
performanceRating <- ggplot(employeeData,aes(PerformanceRating,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("PerformanceRating distribution") + theme
#relationshipSatisfaction <- ggplot(employeeData,aes(RelationshipSatisfaction))+ geom_bar()+ #xlab("") + ylab("Frequency") +ggtitle("Relationship Satisfaction distribution") + theme
workLifeBalance <- ggplot(employeeData,aes(WorkLifeBalance,fill=Attrition))+ geom_bar(position="dodge")+ xlab("") + ylab("Frequency") +ggtitle("Work Life Balance distribution") + theme
grid.arrange(overTime,performanceRating,workLifeBalance,ncol=3)
#grid.arrange(relationshipSatisfaction,workLifeBalance,ncol=2)
```

Using Chi-square test of independence ,independence of categorical variable with *Attrition* can be evaluated.

```{r chi-square,echo=FALSE,warning=FALSE}
dept_chisqr <-chisq.test(table(employeeData$Department,employeeData$Attrition))
bTravel_chisqr <-chisq.test(table(employeeData$BusinessTravel,employeeData$Attrition))
edu_chisqr <-chisq.test(table(employeeData$Education,employeeData$Attrition))
eField_chisqr <-chisq.test(table(employeeData$EducationField,employeeData$Attrition))
esatisfy_chisqr <-chisq.test(table(employeeData$EnvironmentSatisfaction,employeeData$Attrition))
jobinvolve_chisqr <-chisq.test(table(employeeData$JobInvolvement,employeeData$Attrition))
gender_chisqr <-chisq.test(table(employeeData$Gender,employeeData$Attrition))
maritalS_chisqr <-chisq.test(table(employeeData$MaritalStatus,employeeData$Attrition))
jRole_chisqr <-chisq.test(table(employeeData$JobRole,employeeData$Attrition))
jsatisfy_chisqr <-chisq.test(table(employeeData$JobSatisfaction,employeeData$Attrition))
jlevel_chisqr <-chisq.test(table(employeeData$JobLevel,employeeData$Attrition))
stock_chisqr <-chisq.test(table(employeeData$StockOptionLevel,employeeData$Attrition))
oTime_chisqr <- chisq.test(table(employeeData$OverTime,employeeData$Attrition))
prating_chisqr <-chisq.test(table(employeeData$PerformanceRating,employeeData$Attrition))
#relationsat_chisqr <-chisq.test(table(employeeData$RelationshipSatisfaction,
#employeeData$Attrition))
wlbalance_chisqr <-chisq.test(table(employeeData$WorkLifeBalance,employeeData$Attrition))
```

  Variable                |   P- value                     |  
--------------------------|--------------------------------|
  Department              | `r dept_chisqr$p.value `       |
    Business Travel       |  `r bTravel_chisqr$p.value `   |
     Education            | `r edu_chisqr$p.value `        |
    Education Field       |  `r eField_chisqr$p.value `    |
Environment Satisfaction  | `r esatisfy_chisqr$p.value `   |
    Job involvement       |  `r jobinvolve_chisqr$p.value `|
     Gender               | `r gender_chisqr$p.value `     |
    Marital status        |  `r maritalS_chisqr$p.value `  |
     Job role             | `r jRole_chisqr$p.value `      |
    Job Satisfaction      |  `r jsatisfy_chisqr$p.value `  |
    Job level             | `r jlevel_chisqr$p.value `     |
    Stock Options level   |  `r stock_chisqr$p.value `     |
     Overtime             | `r oTime_chisqr$p.value `      |
    Performance Rating    |  `r prating_chisqr$p.value `   |
    Work life balance     |  `r wlbalance_chisqr$p.value ` |
                          |                                |

From the above table, we can infer *Attrition* and *Business Travel* , *Environment Satisfaction*, *Job involvement* , *Marital status* , *Job role* , *Job Satisfaction*, *Job level* , *Stock Options level*,  *Overtime* and *Work life balance* are not independent.


```{r}
CrossTable(employeeData$OverTime,employeeData$Attrition)
```

Attrition rate is higher amoung the employees working over time.Working over time , seams to be a vital indicator. 


##Numeric variables 

```{r,echo=FALSE}
employeeData_numeric <- employeeData[colnames(employeeData) %in% c("Age","DailyRate",
"DistanceFromHome","HourlyRate","MonthlyIncome","MonthlyRate","NumCompaniesWorked",
"PercentSalaryHike","TotalWorkingYears","TrainingTimesLastYear","YearsAtCompany",
"YearsInCurrentRole","YearsSinceLastPromotion","YearsWithCurrManager")]
corrplot(cor(employeeData_numeric),order ="FPC")
```

Correlation plot clearly depicts, high correlation among the following :

* Years with Current Manager, Years at current company and years in current role
* Age and Total working years
* Monthly Income and Total working years
* Years at company and total working years


```{r,echo=FALSE}
percentSalaryHike <-ggplot(employeeData,aes(PercentSalaryHike))+ geom_histogram(binwidth = 2)+ ylab("Frequency")+xlab("")+  xlim(10,30) +ggtitle("Percentage salary hike distribution") +theme
#numCompaniesWorked <-ggplot(employeeData,aes(NumCompaniesWorked))+ geom_histogram(binwidth = 2)+ #xlab("Percentage salary hike") + ylab("Frequency")+  xlim(10,30) +ggtitle("Percentage salary #hike distribution") +theme
grid.arrange(percentSalaryHike,ncol=2)

```

```{r ,echo=FALSE}
totalWorkingYears <-ggplot(employeeData,aes(TotalWorkingYears))+ geom_histogram(binwidth = 3) + ylab("Frequency")+xlab("") +ggtitle("Experience distribution") +theme

YearsAtCompany <-ggplot(employeeData,aes(YearsAtCompany))+ geom_histogram(binwidth = 3) + ylab("Frequency") +xlab("")+ggtitle("Years at company distribution") +theme

yearsInCurrentRole <-ggplot(employeeData,aes(YearsInCurrentRole))+ geom_histogram(binwidth = 2) + ylab("Frequency")+xlab("") + ggtitle("Years in current role distribution") +theme

YearsSinceLastPromotion <-ggplot(employeeData,aes(YearsSinceLastPromotion))+ geom_histogram(binwidth = 5)+ylab("Frequency")+xlab("") +  ggtitle("Years since last promotion distribution") +theme

grid.arrange(totalWorkingYears,YearsAtCompany,ncol=2)
grid.arrange(yearsInCurrentRole,YearsSinceLastPromotion,ncol=2)
```

```{r,echo=FALSE}
trainingTimesLastYear <-ggplot(employeeData,aes(TrainingTimesLastYear))+ geom_histogram(binwidth = 1) +xlab("") +ylab("Frequency") +ggtitle("No. of Trainings since last year distribution") +theme

yearsWithCurrManager <-ggplot(employeeData,aes(YearsWithCurrManager))+ geom_histogram(binwidth = 2)+ ylab("Frequency") +xlab("")+ggtitle("Years with current manager distribution") +theme

grid.arrange(trainingTimesLastYear,yearsWithCurrManager,ncol=2)
```

```{r,echo=FALSE}
dailyRate <-ggplot(employeeData,aes(DailyRate))+ geom_histogram(binwidth = 100)+ ylab("Frequency") +xlab("")+ ggtitle("Daily rate distribution") +theme

hourlyRate <-ggplot(employeeData,aes(HourlyRate))+ geom_histogram(binwidth = 10)+ ylab("Frequency") +xlab("")+ggtitle("Hourly rate distribution") +theme

monthlyIncome <-ggplot(employeeData,aes(MonthlyIncome))+ geom_histogram(binwidth = 1000)+ ylab("Frequency") +xlab("")+ggtitle("Monthly income distribution") +theme

monthlyRate <-ggplot(employeeData,aes(MonthlyRate))+ geom_histogram(binwidth = 1000)+ ylab("Frequency") +xlab("")+ggtitle("Monthly rate distribution") +theme

grid.arrange(dailyRate,hourlyRate,monthlyIncome,monthlyRate, ncol=2,nrow =2)

```


Distribution of daily rate,hourly rate and monthly rate appears to be uniform.Technically they should have a similar distribution as of monthly income.But there appears to be quite a lot of difference.
Since details about the data are not known, these variables will not be considered while modeling. 

## Conclusion

Finally we are left with 16 categorical variables and 11 numeric varaibles.